### 2014-08-05 Studies/2012_Chicago/Report/Chicago_Fire.tex
Chicago: Minor edits

\chapter*{\centering Abstract}
\addcontentsline{toc}{chapter}{Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2 \sfrac{1}{2} story, single family structure in Chicago, IL, that resulted in the death of a fire captain. The inputs for the FDS simulations are documented in this report and are based on the fee scenario including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

@@

On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a Chicago fire captain. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  softwareto provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the creation of a flow path inside the structure.  The document is organized in the following fashion:  Section~\ref{fire_sum} describes the events of the fire incident, Section~\ref{model} describes the relevant input parameters and assumptions used to develop the simulations, Section~\ref{results} presents simulation results, and Section~\ref{discussion} discusses the model results as they relate to firefighter safety and effectiveness.

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident were documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1\textsuperscript{st} floor apartment and worked his way to [the rear of the structure]. He opened the [interior door opening into the rear foyer on the back side of the house] and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2\textsuperscript{nd} floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2\sfrac{1}{2}-inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [the left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [the left side]. A TL39 firefighter went to [the rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door [that connects the 1\textsuperscript{st} floor rear porch to the rear foyer], proceeded to walk in, saw no fire then backed out.

At 1723 hours, the IC radioed the [E123 captain] that there was heavy fire in the covered porch and attic area and that E49 was going to put water on the fire, around the [rear] attic window, but there was no acknowledgment from the [captain]. E49 proceeded to put water on the [rear of the structure]. The IC returned out front to the command post and donned his turnout gear. The lieutenant from TL39, the lieutenant and 2 firefighters from SQ5, and a E123 firefighter/paramedic (FF/PM) were near the kitchen area on the 2\textsuperscript{nd} floor when they heard a loud commotion. The FF/PM heard the [E123 captain] yell ``get out of here.'' The FF/PM had no radio and he couldn't locate the [captain's] radio so he yelled Mayday as he tried to get the [captain] and other crew member untangled. The TL39 Lieutenant and SQ5 firefighters heard the FF/PM's verbal Mayday and the SQ5 Lieutenant tried to transmit a Mayday over heavy radio traffic.

@@

\chapter{Model Description}
\label{model}
The Fire Dynamics Simulator~\cite{FDS_Users_Guide} is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}.  A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

@@

\begin{figure}[!ht]
\centering
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
   \scalebox{1.0}{ \includegraphics[width=3in]{../Figures/exterior_alpha} } &
   \scalebox{1.0}{ \includegraphics[width=2.67in]{../Figures/exterior_charlie} } \\
\end{tabular*}
\caption{Photographs showing the front (left) and rear (right) exterior of the structure after the incident.}
\label{fig:alpha_ex}
\end{figure}
%Fig.~(\ref{fig:smv_interior}) shows 3D Smokeview renderings of cutaways of the beta and delta sides of the structure interior.
%\begin{figure}[!ht]
%\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%\includegraphics[width=.80\textwidth]{../Figures/interior_beta} \\
%\includegraphics[width=.80\textwidth]{../Figures/interior_delta} \\
%\end{tabular*}
%\caption{3D Smokeview rendering of structure interior looking from the charlie side to both the beta side (top) and to the delta side (bottom)}
%\label{fig:smv_interior}
%\end{figure}

@@

Based on a review of these incidents, it is clear that fires with rapidly developing or changing flow paths are a significant hazard to the fire service. The development of or changes in a flow path could be caused by the failure of a component of the structure, such as a door, window, or portion of a ceiling, wall or floor. Environmental conditions such as wind can generate hazardous thermal conditions within a flow path. Uncoordinated ventilation can also be the cause of increased thermal hazards within a flow path.

>>>>

\chapter*{\centering Abstract}
\addcontentsline{toc}{chapter}{Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2 \sfrac{1}{2} story, single family structure in Chicago, IL, that resulted in the death of a fire captain. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

@@

On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a Chicago fire captain. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{fire_sum} provides a summary of the fire incident, Section~\ref{model} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{results} presents the simulation results, and Section~\ref{discussion} discusses the simulation results as they relate to firefighter safety and effectiveness.

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1\textsuperscript{st} floor apartment and worked his way to [the rear of the structure]. He opened the [interior door opening into the rear foyer on the back side of the house] and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2\textsuperscript{nd} floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2\sfrac{1}{2}-inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [the left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [the left side]. A TL39 firefighter went to [the rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door [that connects the 1\textsuperscript{st} floor rear porch to the rear foyer], [proceeded] to walk in, saw no fire then backed out.

At 1723 hours, the IC radioed the [E123 captain] that there was heavy fire in the covered porch and attic area and that E49 was going to put water on the fire, around the [rear] attic window, but there was no acknowledgment from the [captain]. E49 proceeded to put water on the [rear of the structure]. The IC returned out front to the command post and donned his turnout gear. The lieutenant from TL39, the lieutenant and 2 firefighters from SQ5, and a E123 firefighter/paramedic (FF/PM) were near the kitchen area on the 2\textsuperscript{nd} floor when they heard a loud commotion. The FF/PM heard the [E123 captain] yell ``get out of here.'' The FF/PM had no radio and he couldn't locate the [captain's] radio so he yelled Mayday as he tried to get the [captain] and other crew member untangled. The TL39 Lieutenant and SQ5 fire fighters heard the FF/PM's verbal Mayday and the SQ5 Lieutenant tried to transmit a Mayday over heavy radio traffic.

@@

\chapter{Model Description}
\label{model}
Fire Dynamics Simulator~\cite{FDS_Users_Guide} is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}.  A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

@@

\begin{figure}[!ht]
\centering
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
   \scalebox{1.0}{ \includegraphics[width=3in]{../Figures/exterior_alpha} } &
   \scalebox{1.0}{ \includegraphics[width=2.67in]{../Figures/exterior_charlie} } \\
\end{tabular*}
\caption{Photographs showing the front (left) and rear (right) exterior of the structure after the incident.}
\label{fig:alpha_ex}
\end{figure}

%Fig.~(\ref{fig:smv_interior}) shows 3D Smokeview renderings of cutaways of the beta and delta sides of the structure interior.
%\begin{figure}[!ht]
%\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%\includegraphics[width=.80\textwidth]{../Figures/interior_beta} \\
%\includegraphics[width=.80\textwidth]{../Figures/interior_delta} \\
%\end{tabular*}
%\caption{3D Smokeview rendering of structure interior looking from the charlie side to both the beta side (top) and to the delta side (bottom)}
%\label{fig:smv_interior}
%\end{figure}

@@

Based on a review of these incidents, it is clear that fires with rapidly developing or changing flow paths are a significant hazard to the fire service. The development of (or changes to) a flow path could be caused by the failure of a component of the structure, such as a door, window, or portion of a ceiling, wall or floor. Environmental conditions such as wind can generate hazardous thermal conditions within a flow path. Uncoordinated ventilation can also be the cause of increased thermal hazards within a flow path.
